---
title: "Supplements to Analysing Standard Progressive Matrics (SPM-LS) with Bayesian Item Response Models"
author: "Paul Bürkner"
output:
  html_document:
    theme: default
encoding: UTF-8
editor_options: 
  chunk_output_type: console
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

---

# {.tabset}

## Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.width = 10)
```

```{r general_options, cache=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(GGally)
library(brms)
library(mirt)
cores <- 4
iter <- 3000
warmup <- 1000
control <- list(adapt_delta = 0.95)
options(width = 160)
theme_set(theme_default())
```

```{r functions}
scale2 <- function(x, na.rm = TRUE) {
  (x - mean(x, na.rm = na.rm)) / sd(x, na.rm = na.rm)
}

center <- function(x, na.rm = TRUE) {
  x - mean(x, na.rm = na.rm)
}

r2z <- function(r) {
  log((1 + r) / (1 - r)) / 2
}

z2r <- function(z) {
  (exp(2 * z) - 1) / (exp(2 * z) + 1)
}

SW <- suppressWarnings

sum_coding <- function(x, lvls = levels(x)) {
  # codes the first category with -1
  nlvls <- length(lvls)
  stopifnot(nlvls > 1)
  cont <- diag(nlvls)[, -nlvls, drop = FALSE]
  cont[nlvls, ] <- -1
  cont <- cont[c(nlvls, 1:(nlvls - 1)), , drop = FALSE]
  colnames(cont) <- lvls[-1]
  x <- factor(x, levels = lvls)
  contrasts(x) <- cont
  x
}

dummy_coding <- function(x, lvls = levels(x)) {
  x <- factor(x, levels = lvls)
  contrasts(x) <- contr.treatment(levels(x))
  x
}

Sum <- function(df, vars, na.rm = TRUE) {
  # row-wise sum of variables vars stored in df
  vars <- enquo(vars)
  df <- select(df, !!vars)
  all_na <- apply(df, 1, function(x) all(is.na(x)))
  out <- rowSums(select(df, !!vars), na.rm = na.rm)
  ifelse(all_na, NA, out)
}

Mean <- function(df, vars, na.rm = TRUE) {
  # row-wise mean of variables vars stored in df
  vars <- enquo(vars)
  df <- select(df, !!vars)
  all_na <- apply(df, 1, function(x) all(is.na(x)))
  out <- rowMeans(select(df, !!vars), na.rm = na.rm)
  ifelse(all_na, NA, out)
}
```

## Data Preparation

```{r}
answers <- c(7, 6, 8, 2, 1, 5, 1, 6, 3, 2, 4, 5)
names(answers) <- paste0("SPM", seq_along(answers))
```

```{r}
spm_long <- read.csv("data/dataset.csv")  %>%
  mutate(person = seq_len(n())) %>%
  gather("item", "response", SPM1:SPM12) %>%
  mutate(
    answer = answers[item],
    response2 = as.numeric(response == answer),
    item = as.numeric(sub("^SPM", "", item))
  )
```

```{r}
spm_wide <- spm_long %>% 
  select(person, item, response2) %>%
  spread("item", "response2") %>%
  select(-person)
```

## Analysis {.tabset}

### 1PL Models {.tabset}

#### Model Fitting

```{r}
formula_1pl <- bf(
  formula = response2 ~ 1 + (1 | item) + (1 | person),
  family = brmsfamily("bernoulli", link = "logit")
)

prior_1pl <- 
  prior("normal(0, 2)", class = "Intercept")
  prior("normal(0, 3)", class = "sd", group = "person") + 
  prior("normal(0, 3)", class = "sd", group = "item")

fit_1pl <- brm(
  formula = formula_1pl,
  data = spm_long,
  prior = prior_1pl,
  chains = cores, iter = iter, warmup = warmup,
  cores = cores, control = control,
  file = "models/fit_1pl"
)

fit_1pl <- add_loo(fit_1pl)
```

```{r}
# obtain basic summaries
summary(fit_1pl)
plot(fit_1pl)
```

```{r, results="hide"}
mirt_1pl <- mirt(spm_wide, model = 1, itemtype = "Rasch", SE = TRUE,
                 technical = list(NCYCLES = 10000))
```

#### Person Parameters

```{r}
# extract person parameters
person_pars_1pl <- 
  ranef(fit_1pl, summary = FALSE)$person[, , "Intercept"] 

person_sds_1pl <- apply(person_pars_1pl, 1, sd)

person_pars_1pl <- person_pars_1pl %>%
  sweep(1, person_sds_1pl, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column(var = "person") %>%
  mutate(person = as.numeric(person))

# plot person parameters
person_pars_1pl %>%
  arrange(Estimate) %>%
  mutate(id2 = seq_len(n())) %>%
	ggplot(aes(id2, Estimate, ymin = Q2.5, ymax = Q97.5)) +
	geom_pointrange(alpha = 0.7) +
	coord_flip() +
	labs(x = "Person Number (sorted after Estimate)") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
mirt_person_pars_1pl <- fscores(mirt_1pl, full.scores.SE = TRUE) %>%
  as.data.frame() %>%
  rename(Estimate = "F1", Est.Error = "SE_F1") %>%
  mutate(
    # person parameters are not scaled appropriately by mirt for the 1PL model
    scale = sd(Estimate) / sd(person_pars_1pl$Estimate),
    Estimate = Estimate / scale,
    Est.Error = Est.Error / scale,
    Q2.5 = Estimate - 1.96 * Est.Error, 
    Q97.5 = Estimate + 1.96 * Est.Error,
    person = seq_len(n())
  ) %>%
  select(-scale)
```

```{r}
# plot comparison of person parameters
person_pars_1pl_all <- person_pars_1pl %>% 
  bind_cols(mirt_person_pars_1pl) %>%
  mutate(
    UIW = Q97.5 - Q2.5,
    UIW1 = Q97.51 - Q2.51
  )

person_pars_1pl_all %>%
	ggplot(aes(Estimate, Estimate1, color = UIW)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(-3, 2), y = c(-3, 2)) +
	labs(
		x = "Estimate (brms)",
	  y = "Estimate (mirt)",
		color = "UIW (brms)"
	)

person_pars_1pl_all %>%
	ggplot(aes(UIW, UIW1, color = Estimate)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(1, 2.4), y = c(1, 2.4)) +
	labs(
		x = "UIW (brms)",
	  y = "UIW (mirt)",
		color = "Estimate (brms)"
	)
```

#### Item Parameters

```{r, fig.height=4}
# extract item parameters
item_pars_1pl <- coef(fit_1pl, summary = FALSE)$item[, , "Intercept"] %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column() %>%
	rename(item = "rowname") %>%
	mutate(
	  item = as.numeric(item),
	  package = "brms"
	)
```

```{r}
mirt_item_pars_1pl <- coef(mirt_1pl, as.data.frame = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  filter(str_detect(rowname, "\\.d$")) %>%
  mutate(
    item = str_match(rowname, "^X[[:digit:]]+") %>%
      str_remove("^X") %>%
      as.numeric(),
    item = item - 0.3,
    package = "mirt"
  ) %>%
  rename(Estimate = par, Q2.5 = CI_2.5, Q97.5 = CI_97.5)
```
  
```{r}
# plot item parameters
item_pars_1pl %>%
  bind_rows(mirt_item_pars_1pl) %>%
	ggplot(aes(item, Estimate, ymin = Q2.5, ymax = Q97.5, color = package)) +
  scale_x_continuous(breaks = 1:12) +
	geom_pointrange() +
  scale_color_manual(name = "Package", values = c("black", "grey")) +
	coord_flip() +
	labs(x = "Item Number")
```

### 2PL Models {.tabset}

#### Model Fitting

```{r}
formula_2pl <- bf(
  response2 ~ beta + exp(logalpha) * theta,
  nl = TRUE,
  theta ~ 0 + (1 | person),
  beta ~ 1 + (1 |i| item),
  logalpha ~ 1 + (1 |i| item),
  family = brmsfamily("bernoulli", link = "logit")
)

prior_2pl <- 
  prior("normal(0, 2)", class = "b", nlpar = "beta") +
  prior("normal(0, 1)", class = "b", nlpar = "logalpha") +
  prior("normal(0, 1)", class = "sd", group = "person", nlpar = "theta") + 
  prior("normal(0, 3)", class = "sd", group = "item", nlpar = "beta") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logalpha")

fit_2pl <- brm(
  formula = formula_2pl,
  data = spm_long,
  prior = prior_2pl,
  chains = cores, iter = iter, warmup = warmup,
  cores = cores, control = control, inits = 0,
  file = "models/fit_2pl"
)

fit_2pl <- add_loo(fit_2pl)
```

```{r}
# obtain some basic summaries
summary(fit_2pl)
plot(fit_2pl, pars = c("sd", "cor"))
```

```{r, results="hide"}
mirt_2pl <- mirt(spm_wide, model = 1, itemtype = "2PL", SE = TRUE,
                 technical = list(NCYCLES = 10000))
```

#### Person Parameters

```{r}
# extract and scale person parameters
person_pars_2pl <- 
  ranef(fit_2pl, summary = FALSE)$person[, , "theta_Intercept"]

person_sds_2pl <- apply(person_pars_2pl, 1, sd)

person_pars_2pl <- person_pars_2pl %>%
  sweep(1, person_sds_2pl, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column(var = "person") %>%
  mutate(person = as.numeric(person))

# plot person parameters
person_pars_2pl %>%
  arrange(Estimate) %>%
  mutate(id2 = seq_len(n())) %>%
	ggplot(aes(id2, Estimate, ymin = Q2.5, ymax = Q97.5)) +
	geom_pointrange(alpha = 0.7) +
	coord_flip() +
	labs(x = "Person Number (sorted after Estimate)") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
mirt_person_pars_2pl <- fscores(mirt_2pl, full.scores.SE = TRUE) %>%
  as.data.frame() %>%
  rename(Estimate = "F1", Est.Error = "SE_F1") %>%
  mutate(
    Q2.5 = Estimate - 1.96 * Est.Error, 
    Q97.5 = Estimate + 1.96 * Est.Error,
    person = seq_len(n())
  )
```

```{r}
# plot comparison of person parameters
person_pars_2pl_all <- person_pars_2pl %>% 
  bind_cols(mirt_person_pars_2pl) %>%
  mutate(
    UIW = Q97.5 - Q2.5,
    UIW1 = Q97.51 - Q2.51
  )

person_pars_2pl_all %>%
	ggplot(aes(Estimate, Estimate1, color = UIW)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(-3, 2), y = c(-3, 2)) +
	labs(
		x = "Estimate (brms)",
	  y = "Estimate (mirt)",
		color = "UIW (brms)"
	)

person_pars_2pl_all %>%
	ggplot(aes(UIW, UIW1, color = Estimate)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(1, 2.4), y = c(1, 2.4)) +
	labs(
		x = "UIW (brms)",
	  y = "UIW (mirt)",
		color = "Estimate (brms)"
	)
```

#### Item Parameters

```{r, fig.height=4}
# extract item parameters
item_pars_2pl <- coef(fit_2pl, summary = FALSE)$item

# plot item parameters
# locations
beta <- item_pars_2pl[, , "beta_Intercept"] %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# slopes
alpha <- item_pars_2pl[, , "logalpha_Intercept"] %>%
  exp() %>%
  sweep(1, person_sds_2pl, "*") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

item_pars_2pl <- bind_rows(beta, alpha, .id = "nlpar") %>%
	rename(item = "rowname") %>%
	mutate(item = as.numeric(item)) %>%
	mutate(
	  nlpar = factor(nlpar, labels = c("Location", "Slope")),
	  package = "brms"  
	)
```

```{r, results="hide"}
mirt_item_pars_2pl <- coef(mirt_2pl, as.data.frame = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  filter(str_detect(rowname, "\\.(a1|d)$")) %>%
  mutate(
    item = str_match(rowname, "^X[[:digit:]]+") %>%
      str_remove("^X") %>%
      as.numeric(),
    item = item - 0.3,
    nlpar = str_match(rowname, "\\.[^\\.]+$") %>%
      str_remove("^\\.") %>%
      factor(
        levels = c("d", "a1"),
        labels = c("Location", "Slope")
      ),
    package = "mirt"
  ) %>%
  rename(Estimate = par, Q2.5 = CI_2.5, Q97.5 = CI_97.5) 
```

```{r}
# plot locations and slope next to each other
item_pars_2pl %>%
  bind_rows(mirt_item_pars_2pl) %>%
	ggplot(aes(item, Estimate, ymin = Q2.5, ymax = Q97.5, color = package)) +
  scale_x_continuous(breaks = 1:12) +
	facet_wrap("nlpar", scales = "free_x") +
  geom_pointrange() +
  scale_color_manual(name = "Package", values = c("black", "grey")) +
	coord_flip() +
	labs(x = "Item Number")
```

### 3PL models with fixed guessing probability {.tabset}

#### Model Fitting

```{r}
formula_3pl_fixed <- bf(
  response2 ~ 0.125 + 0.875 * inv_logit(beta + exp(logalpha) * theta),
  nl = TRUE,
  theta ~ 0 + (1 | person),
  beta ~ 1 + (1 |i| item),
  logalpha ~ 1 + (1 |i| item),
  family = brmsfamily("bernoulli", link = "identity")
)

prior_3pl_fixed <- 
  prior("normal(0, 2)", class = "b", nlpar = "beta") +
  prior("normal(0, 1)", class = "b", nlpar = "logalpha") +
  prior("normal(0, 1)", class = "sd", group = "person", nlpar = "theta") + 
  prior("normal(0, 3)", class = "sd", group = "item", nlpar = "beta") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logalpha")

fit_3pl_fixed <- brm(
  formula = formula_3pl_fixed,
  data = spm_long,
  prior = prior_3pl_fixed,
  chains = cores, iter = iter, warmup = warmup,
  cores = cores, control = control, inits = 0,
  file = "models/fit_3pl_fixed"
)

fit_3pl_fixed <- add_loo(fit_3pl_fixed)
```

```{r}
summary(fit_3pl_fixed)
```

```{r}
mirt_3pl_fixed <- mirt(
  spm_wide, model = 1, itemtype = "2PL", guess = 0.11, 
  SE = TRUE, technical = list(NCYCLES = 10000)
)
```

#### Person Parameters

```{r}
# extract and scale person parameters
person_pars_3pl_fixed <- 
  ranef(fit_3pl_fixed, summary = FALSE)$person[, , "theta_Intercept"]

person_sds_3pl_fixed <- apply(person_pars_3pl_fixed, 1, sd)

person_pars_3pl_fixed <- person_pars_3pl_fixed %>%
  sweep(1, person_sds_3pl_fixed, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column(var = "person") %>%
  mutate(person = as.numeric(person))

# plot person parameters
person_pars_3pl_fixed %>%
  arrange(Estimate) %>%
  mutate(id2 = seq_len(n())) %>%
	ggplot(aes(id2, Estimate, ymin = Q2.5, ymax = Q97.5)) +
	geom_pointrange(alpha = 0.7) +
	coord_flip() +
	labs(x = "Person Number (sorted after Estimate)") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
mirt_person_pars_3pl_fixed <- 
  fscores(mirt_3pl_fixed, full.scores.SE = TRUE) %>%
  as.data.frame() %>%
  rename(Estimate = "F1", Est.Error = "SE_F1") %>%
  mutate(
    Q2.5 = Estimate - 1.96 * Est.Error, 
    Q97.5 = Estimate + 1.96 * Est.Error,
    person = seq_len(n())
  )
```

```{r}
# plot comparison of person parameters
person_pars_3pl_fixed_all <- person_pars_3pl_fixed %>% 
  bind_cols(mirt_person_pars_3pl_fixed) %>%
  mutate(
    UIW = Q97.5 - Q2.5,
    UIW1 = Q97.51 - Q2.51
  )

person_pars_3pl_fixed_all %>%
	ggplot(aes(Estimate, Estimate1, color = UIW)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(-3, 2), y = c(-3, 2)) +
	labs(
		x = "Estimate (brms)",
	  y = "Estimate (mirt)",
		color = "UIW (brms)"
	)

person_pars_3pl_fixed_all %>%
	ggplot(aes(UIW, UIW1, color = Estimate)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(1, 2.4), y = c(1, 2.4)) +
	labs(
		x = "UIW (brms)",
	  y = "UIW (mirt)",
		color = "Estimate (brms)"
	)
```

#### Item Parameters

```{r, fig.height=4}
# extract item parameters
item_pars_3pl_fixed <- coef(fit_3pl_fixed, summary = FALSE)$item

# locations
beta <- item_pars_3pl_fixed[, , "beta_Intercept"] %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# slopes
alpha <- item_pars_3pl_fixed[, , "logalpha_Intercept"] %>%
  exp() %>%
  sweep(1, person_sds_3pl_fixed, "*") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

item_pars_3pl_fixed <- bind_rows(beta, alpha, .id = "nlpar") %>%
	rename(item = "rowname") %>%
	mutate(item = as.numeric(item)) %>%
	mutate(
	  nlpar = factor(nlpar, labels = c("Location", "Slope")),
	  package = "brms"  
	)
```

```{r, results="hide"}
mirt_item_pars_3pl_fixed <- coef(mirt_3pl_fixed, as.data.frame = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  filter(str_detect(rowname, "\\.(a1|d)$")) %>%
  mutate(
    item = str_match(rowname, "^X[[:digit:]]+") %>%
      str_remove("^X") %>%
      as.numeric(),
    item = item - 0.3,
    nlpar = str_match(rowname, "\\.[^\\.]+$") %>%
      str_remove("^\\.") %>%
      factor(
        levels = c("d", "a1"),
        labels = c("Location", "Slope")
      ),
    package = "mirt"
  ) %>%
  rename(Estimate = par, Q2.5 = CI_2.5, Q97.5 = CI_97.5) 
```

```{r}
# plot locations and slope next to each other
item_pars_3pl_fixed %>%
  bind_rows(mirt_item_pars_3pl_fixed) %>%
	ggplot(aes(item, Estimate, ymin = Q2.5, ymax = Q97.5, color = package)) +
  scale_x_continuous(breaks = 1:12) +
	facet_wrap("nlpar", scales = "free_x") +
	geom_pointrange() +
	scale_color_manual(name = "Package", values = c("black", "grey")) +
	coord_flip() +
	labs(x = "Item Number")
```

### 3PL models {.tabset}

#### Model Fitting

```{r}
formula_3pl <- bf(
  response2 ~ gamma + (1 - gamma) * 
    inv_logit(beta + exp(logalpha) * theta),
  nl = TRUE,
  theta ~ 0 + (1 | person),
  beta ~ 1 + (1 |i| item),
  logalpha ~ 1 + (1 |i| item),
  logitgamma ~ 1 + (1 |i| item),
  nlf(gamma ~ inv_logit(logitgamma)),
  family = brmsfamily("bernoulli", link = "identity")
)

prior_3pl <- 
  prior("normal(0, 2)", class = "b", nlpar = "beta") +
  prior("normal(0, 1)", class = "b", nlpar = "logalpha") +
  prior("normal(-2, 0.5)", class = "b", nlpar = "logitgamma") +
  prior("normal(0, 1)", class = "sd", group = "person", nlpar = "theta") + 
  prior("normal(0, 3)", class = "sd", group = "item", nlpar = "beta") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logalpha") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logitgamma")

fit_3pl <- brm(
  formula = formula_3pl,
  data = spm_long,
  prior = prior_3pl,
  chains = cores, iter = iter, warmup = warmup,
  cores = cores, control = control, inits = 0,
  file = "models/fit_3pl"
)

fit_3pl <- add_loo(fit_3pl)
```

```{r}
summary(fit_3pl)
```

```{r, results = "hide"}
mirt_3pl <- mirt(spm_wide, model = 1, itemtype = "3PL", SE = TRUE,
                 technical = list(NCYCLES = 10000))
```

#### Person Parameters

```{r}
# extract and scale person parameters
person_pars_3pl <- 
  ranef(fit_3pl, summary = FALSE)$person[, , "theta_Intercept"]

person_sds_3pl <- apply(person_pars_3pl, 1, sd)

person_pars_3pl <- person_pars_3pl %>%
  sweep(1, person_sds_3pl, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column(var = "person") %>%
  mutate(person = as.numeric(person))

# plot person parameters
person_pars_3pl %>%
  arrange(Estimate) %>%
  mutate(id2 = seq_len(n())) %>%
	ggplot(aes(id2, Estimate, ymin = Q2.5, ymax = Q97.5)) +
	geom_pointrange(alpha = 0.7) +
	coord_flip() +
	labs(x = "Person Number (sorted after Estimate)") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
mirt_person_pars_3pl <- fscores(mirt_3pl, full.scores.SE = TRUE) %>%
  as.data.frame() %>%
  rename(Estimate = "F1", Est.Error = "SE_F1") %>%
  mutate(
    Q2.5 = Estimate - 1.96 * Est.Error, 
    Q97.5 = Estimate + 1.96 * Est.Error,
    person = seq_len(n())
  )
```

```{r}
# plot comparison of person parameters
person_pars_3pl_all <- person_pars_3pl %>% 
  bind_cols(mirt_person_pars_3pl) %>%
  mutate(
    UIW = Q97.5 - Q2.5,
    UIW1 = Q97.51 - Q2.51
  )

person_pars_3pl_all %>%
	ggplot(aes(Estimate, Estimate1, color = UIW)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(-3, 2), y = c(-3, 2)) +
	labs(
		x = "Estimate (brms)",
	  y = "Estimate (mirt)",
		color = "UIW (brms)"
	)

person_pars_3pl_all %>%
	ggplot(aes(UIW, UIW1, color = Estimate)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(1, 2.4), y = c(1, 2.4)) +
	labs(
		x = "UIW (brms)",
	  y = "UIW (mirt)",
		color = "Estimate (brms)"
	)
```

#### Item Parameters

```{r, fig.height=4}
# extract item parameters
item_pars_3pl <- coef(fit_3pl, summary = FALSE)$item

# plot item parameters
# locations
beta <- item_pars_3pl[, , "beta_Intercept"] %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# slopes
alpha <- item_pars_3pl[, , "logalpha_Intercept"] %>%
	exp() %>%
  sweep(1, person_sds_3pl, "*") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# guessing parameters
gamma <- item_pars_3pl[, , "logitgamma_Intercept"] %>%
	inv_logit_scaled() %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

item_pars_3pl <- bind_rows(beta, alpha, gamma, .id = "nlpar") %>%
	rename(item = "rowname") %>%
	mutate(item = as.numeric(item)) %>%
	mutate(
	  nlpar = factor(nlpar, labels = c("Location", "Slope", "Guessing")),
	  package = "brms"  
	)
```

```{r, results="hide"}
mirt_item_pars_3pl <- coef(mirt_3pl, as.data.frame = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  filter(str_detect(rowname, "\\.(a1|d|g)$")) %>%
  mutate(
    item = str_match(rowname, "^X[[:digit:]]+") %>%
      str_remove("^X") %>%
      as.numeric(),
    item = item - 0.3,
    nlpar = str_match(rowname, "\\.[^\\.]+$") %>%
      str_remove("^\\.") %>%
      factor(
        levels = c("d", "a1", "g"),
        labels = c("Location", "Slope", "Guessing")
      ),
    package = "mirt"
  ) %>%
  rename(Estimate = par, Q2.5 = CI_2.5, Q97.5 = CI_97.5) 
```

```{r}
item_pars_3pl %>%
  bind_rows(mirt_item_pars_3pl) %>%
  ggplot(aes(item, Estimate, ymin = Q2.5, ymax = Q97.5, color = package)) +
  scale_x_continuous(breaks = 1:12) +
	facet_wrap("nlpar", scales = "free_x") +
	geom_pointrange() +
  scale_color_manual(name = "Package", values = c("black", "grey")) +
	coord_flip() +
	labs(x = "Item Number")
```

### 4PL models {.tabset}

#### Model Fitting

```{r}
formula_4pl <- bf(
  response2 ~ gamma + (1 - gamma - psi) * 
    inv_logit(beta + exp(logalpha) * theta),
  nl = TRUE,
  theta ~ 0 + (1 | person),
  beta ~ 1 + (1 |i| item),
  logalpha ~ 1 + (1 |i| item),
  logitgamma ~ 1 + (1 |i| item),
  nlf(gamma ~ inv_logit(logitgamma)),
  logitpsi ~ 1 + (1 |i| item),
  nlf(psi ~ inv_logit(logitpsi)),
  family = brmsfamily("bernoulli", link = "identity")
)

prior_4pl <- 
  prior("normal(0, 2)", class = "b", nlpar = "beta") +
  prior("normal(0, 1)", class = "b", nlpar = "logalpha") +
  prior("normal(-2, 0.5)", class = "b", nlpar = "logitgamma") +
  prior("normal(-2, 0.5)", class = "b", nlpar = "logitpsi") +
  prior("normal(0, 1)", class = "sd", group = "person", nlpar = "theta") + 
  prior("normal(0, 3)", class = "sd", group = "item", nlpar = "beta") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logalpha") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logitgamma") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logitpsi")

fit_4pl <- brm(
  formula = formula_4pl,
  data = spm_long,
  prior = prior_4pl,
  chains = cores, iter = iter, warmup = warmup,
  cores = cores, control = control, inits = 0,
  file = "models/fit_4pl"
)
fit_4pl <- add_loo(fit_4pl)
```

```{r}
summary(fit_4pl)
```

```{r, results="hide"}
mirt_4pl <- mirt(
  spm_wide, model = 1, itemtype = "4PL", SE = TRUE,
  technical = list(NCYCLES = 10000)
)
```

#### Person Parameters

```{r}
# extract and scale person parameters
person_pars_4pl <- 
  ranef(fit_4pl, summary = FALSE)$person[, , "theta_Intercept"]

person_sds_4pl <- apply(person_pars_4pl, 1, sd)

person_pars_4pl <- person_pars_4pl %>%
  sweep(1, person_sds_4pl, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column(var = "person") %>%
  mutate(person = as.numeric(person))

# plot person parameters
person_pars_4pl %>%
  arrange(Estimate) %>%
  mutate(id2 = seq_len(n())) %>%
	ggplot(aes(id2, Estimate, ymin = Q2.5, ymax = Q97.5)) +
	geom_pointrange(alpha = 0.7) +
	coord_flip() +
	labs(x = "Person Number (sorted after Estimate)") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
mirt_person_pars_4pl <- fscores(mirt_4pl, full.scores.SE = TRUE) %>%
  as.data.frame() %>%
  rename(Estimate = "F1", Est.Error = "SE_F1") %>%
  mutate(
    Q2.5 = Estimate - 1.96 * Est.Error, 
    Q97.5 = Estimate + 1.96 * Est.Error,
    person = seq_len(n())
  )
```

```{r}
# plot comparison of person parameters
person_pars_4pl_all <- person_pars_4pl %>% 
  bind_cols(mirt_person_pars_4pl) %>%
  mutate(
    UIW = Q97.5 - Q2.5,
    UIW1 = Q97.51 - Q2.51
  )

person_pars_4pl_all %>%
	ggplot(aes(Estimate, Estimate1, color = UIW)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(-3, 2), y = c(-3, 2)) +
	labs(
		x = "Estimate (brms)",
	  y = "Estimate (mirt)",
		color = "UIW (brms)"
	)

person_pars_4pl_all %>%
	ggplot(aes(UIW, UIW1, color = Estimate)) +
	geom_point() +
	geom_abline() +
	scale_color_viridis_c() +
	lims(x = c(1, 2.4), y = c(1, 2.4)) +
	labs(
		x = "UIW (brms)",
	  y = "UIW (mirt)",
		color = "Estimate (brms)"
	)
```

#### Item Parameters

```{r}
# extract item parameters
item_pars_4pl <- coef(fit_4pl, summary = FALSE)$item

# locations
beta <- item_pars_4pl[, , "beta_Intercept"] %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# slopes
alpha <- item_pars_4pl[, , "logalpha_Intercept"] %>%
	exp() %>%
  sweep(1, person_sds_4pl, "*") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# guessing parameters
gamma <- item_pars_4pl[, , "logitgamma_Intercept"] %>%
	inv_logit_scaled() %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# lapse parameters
psi <- item_pars_4pl[, , "logitpsi_Intercept"] %>%
	inv_logit_scaled() %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

item_pars_4pl <- bind_rows(beta, alpha, gamma, psi, .id = "nlpar") %>%
	rename(item = "rowname") %>%
	mutate(item = as.numeric(item)) %>%
	mutate(
	  nlpar = nlpar %>% 
	    factor(labels = c("Location", "Slope", "Guessing", "Lapse")),
	  package = "brms"
	)
```

```{r, results="hide"}
mirt_item_pars_4pl <- coef(mirt_4pl, as.data.frame = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  filter(str_detect(rowname, "\\.(a1|d|g|u)$")) %>%
  mutate(
    item = str_match(rowname, "^X[[:digit:]]+") %>%
      str_remove("^X") %>%
      as.numeric(),
    item = item - 0.3,
    nlpar = str_match(rowname, "\\.[^\\.]+$") %>%
      str_remove("^\\.") %>%
      factor(
        levels = c("d", "a1", "g", "u"),
        labels = c("Location", "Slope", "Guessing", "Lapse")
      ),
    # invert guessing parameter
    par = ifelse(nlpar == "Lapse", 1 - par, par),
    # cannot compute CIs as inverting the information matrix fails
    CI_2.5 = par,
    CI_97.5 = par,
    package = "mirt"
  ) %>%
  rename(Estimate = par, Q2.5 = CI_2.5, Q97.5 = CI_97.5)
```

```{r, fig.height=8}
item_pars_4pl %>%
  bind_rows(mirt_item_pars_4pl) %>%
	ggplot(aes(item, Estimate, ymin = Q2.5, ymax = Q97.5, color = package)) +
  scale_x_continuous(breaks = 1:12) +
	facet_wrap("nlpar", scales = "free_x") +
	geom_pointrange() +
  scale_color_manual(name = "Package", values = c("black", "grey")) +
	coord_flip() +
	labs(x = "Item Number")
```

### Comparison of models

#### Out-Of-Sample Predictive Performance

```{r}
loo_compare(fit_1pl, fit_2pl, fit_3pl_fixed, fit_3pl, fit_4pl)
```

#### Person Parameters

```{r}
person_pars_all <- bind_rows(
  "1PL: brms" = person_pars_1pl,
  "1PL: mirt" = mirt_person_pars_1pl, 
  "2PL: brms" = person_pars_2pl, 
  "2PL: mirt" = mirt_person_pars_2pl, 
  "3PL(f): brms" = person_pars_3pl_fixed,
  "3PL(f): mirt" = mirt_person_pars_3pl_fixed,
  "3PL: brms" = person_pars_3pl,
  "3PL: mirt" = mirt_person_pars_3pl,
  "4PL: brms" = person_pars_4pl, 
  "4PL: mirt" = mirt_person_pars_4pl, 
  .id = "model"
) %>%
  mutate(model = factor(model, levels = unique(model)))
```

```{r, fig.height=12, fig.width=12}
person_pars_all %>%
  select(Estimate, model, person) %>% 
  spread(model, Estimate) %>%
  select(-person) %>%
  GGally::ggpairs(aes(alpha = 0.4))
```

```{r, fig.height=12, fig.width=12}
person_pars_all %>%
  select(Est.Error, model, person) %>% 
  spread(model, Est.Error) %>%
  select(-person) %>%
  GGally::ggpairs(aes(alpha = 0.4))
```

### 4PL models in Conditional Formulation {.tabset}

#### Model Fitting

```{r}
formula_4pl_cond <- bf(
  response2 ~ sigma * zeta + (1 - sigma) * inv_logit(beta + exp(logalpha) * theta),
  nl = TRUE,
  theta ~ 0 + (1 | person),
  beta ~ 1 + (1 |i| item),
  logalpha ~ 1 + (1 |i| item),
  logitsigma ~ 1 + (1 |i| item),
  nlf(sigma ~ inv_logit(logitsigma)),
  logitzeta ~ 1 + (1 |i| item),
  nlf(zeta ~ inv_logit(logitzeta)),
  family = brmsfamily("bernoulli", link = "identity")
)

prior_4pl_cond <- 
  prior("normal(0, 2)", class = "b", nlpar = "beta") +
  prior("normal(0, 1)", class = "b", nlpar = "logalpha") +
  prior("normal(-2, 0.5)", class = "b", nlpar = "logitsigma") +
  prior("normal(-2, 0.5)", class = "b", nlpar = "logitzeta") +
  prior("normal(0, 1)", class = "sd", group = "person", nlpar = "theta") + 
  prior("normal(0, 3)", class = "sd", group = "item", nlpar = "beta") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logalpha") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logitsigma") +
  prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logitzeta")

fit_4pl_cond <- brm(
  formula = formula_4pl_cond,
  data = spm_long,
  prior = prior_4pl_cond,
  chains = cores, iter = iter, warmup = warmup,
  cores = cores, control = control, inits = 0,
  file = "models/fit_4pl_cond"
)

fit_4pl_cond <- add_loo(fit_4pl_cond)
```

```{r}
summary(fit_4pl_cond)
```

#### Person Parameters

```{r}
# extract and scale person parameters
person_pars_4pl_cond <- 
  ranef(fit_4pl_cond, summary = FALSE)$person[, , "theta_Intercept"]

person_sds_4pl_cond <- apply(person_pars_4pl_cond, 1, sd)

person_pars_4pl_cond <- person_pars_4pl_cond %>%
  sweep(1, person_sds_4pl_cond, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column(var = "person") %>%
  mutate(person = as.numeric(person))

# plot person parameters
person_pars_4pl_cond %>%
  arrange(Estimate) %>%
  mutate(id2 = seq_len(n())) %>%
	ggplot(aes(id2, Estimate, ymin = Q2.5, ymax = Q97.5)) +
	geom_pointrange(alpha = 0.7) +
	coord_flip() +
	labs(x = "Person Number (sorted after Estimate)") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

#### Item Parameters

```{r, fig.height=8}
# extract item parameters
item_pars_4pl_cond <- coef(fit_4pl_cond, summary = FALSE)$item

# locations
beta <- item_pars_4pl_cond[, , "beta_Intercept"] %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# slopes
alpha <- item_pars_4pl_cond[, , "logalpha_Intercept"] %>%
	exp() %>%
  sweep(1, person_sds_4pl_cond, "/") %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# false answers probability
sigma <- item_pars_4pl_cond[, , "logitsigma_Intercept"] %>%
	inv_logit_scaled()

# conditional guessing probability
zeta <- item_pars_4pl_cond[, , "logitzeta_Intercept"] %>%
	inv_logit_scaled()

# guessing parameters
gamma <- (sigma * zeta) %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# lapse parameters
psi <- (sigma * (1 - zeta)) %>%
  posterior_summary() %>%
	as_tibble() %>%
	rownames_to_column()

# plot locations and slope next to each other
bind_rows(beta, alpha, gamma, psi, .id = "nlpar") %>%
	rename(item = "rowname") %>%
	mutate(item = as.numeric(item)) %>%
	mutate(nlpar = factor(
	  nlpar, labels = c("Location", "Slope", "Guessing", "Lapse")
	)) %>%
	ggplot(aes(item, Estimate, ymin = Q2.5, ymax = Q97.5)) +
  scale_x_continuous(breaks = 1:12) +
	facet_wrap("nlpar", scales = "free_x") +
	geom_pointrange() +
	coord_flip() +
	labs(x = "Item Number")
```
